---
title: "2020-03-19-customRthemes"
author: "Rayna M Harris"
date: "2/17/2020"
output: html_document
---

# Creating customizing colors palette and themes for ggplot

This is 15 minute R exercise or tutorial for making custom color pallets and themes to give a consistent look to ggplots. 


```{r setup, echo = F, include=F}
library(tidyverse)
library(cowplot)
knitr::opts_chunk$set(echo = TRUE, fig.path = "./images/")
```

## A basic ggplot without color

I consult the [R for Data Science](https://r4ds.had.co.nz/) book a lot. This blog post starts with [the code used in Chapter 3](https://r4ds.had.co.nz/data-visualisation.html). I like using this `mpg` dataset because it comes with the tidyverse installation, so I don't need to find an external source of data. 

```{r customRthemes1}
## start with basic plot
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Two gglots using the same colors for different factors

When we use the default `ggplot` color palette for four or five different plots in a document, different factors will take on the same color. This deminished the value of using color to convey information.  

```{r customRthemes2}
## color by factors, default
a <- ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = as.factor(cyl)))
b <-  ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
plot_grid(a,b)
```

## Two ggplots with custom and unique color palletes

Before assigning colors, I arrage the factors in a specific order.

```{r }
## specify order
levelscyl <- c("4", "5", "6",  "8" )
levelsclass <- c("2seater", "subcompact", "compact", "midsize",
                 "suv", "minivan", "pickup")
mpg$cyl <- factor(mpg$cyl, levels = levelscyl)
mpg$class <- factor(mpg$class, levels = levelsclass)
```

I use [https://colorbrewer2.org/](https://colorbrewer2.org/) to pick  colorblind and photocopy safe colors and get their hex codes. Then I assign the hex codes to factors by name. Finally, I group all my colors into one string that I name `colorsall` for easy use. 

```{r}

## pick colors using http://colorbrewer2.org/
colorscyl <- c("4" = "#cbc9e2", 
               "5" = "#9e9ac8", 
               "6" = "#756bb1",
               "8" = "#54278f")

colorsclass <- c("2seater" = "#e41a1c", 
                 "subcompact" = "#1b7837", 
                 "compact" = "#7fbf7b", "midsize" = "#d9f0d3",
                 "suv" = "#d8b365", 
                 "minivan" = "#dfc27d", "pickup" = "#bf812d")

# combine all color palletts into one for quick and easy use
colorsall <- c(colorscyl, colorsclass)
```

Now, I can add the colors to my plot with `scale_color_manual(values = colorsall)`.

```{r customRthemes3}
c <- ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cyl)) +
  scale_color_manual(values = colorsall)
d <- ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) +
    scale_color_manual(values = colorsall)

plot_grid(c,d)
```

By storing the color in a string outside the plot, it's much easier to update one string and change the colors of all your figures than to change the color pallete for each figure indivdually. 

## ggplot with a custom theme

Finally, I like to create custom themes that save you from copying and pasting the same line of code many many time. In this cause we can use `+ mytheme()` to add four lines of code to given figure in order to rotate the x-axis text, make the y-axis font face bold, and ajust the font size to 11 and use the classic ggplot theme. 

```{r customRthemes4}
mytheme <- function(){
  theme_classic(base_size = 11) +
    theme(legend.position = "bottom",
          axis.title.y = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1))
}

e <- ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = cyl)) +
  scale_color_manual(values = colorsall)  +
  mytheme() 

f <- ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) +
    scale_color_manual(values = colorsall)  +
  mytheme() 

plot_grid(e,f, align = "hv", rel_widths = c(0.4, 0.6))
```

Now it's your to experiment with the code. Try these:

- [ ] change the plot to a box plot with `geom_boxplot`
- [ ] experiment with `scale_color_manual` and `scale_color_manual`
- [ ] change the colors of a factor in the `diamond` dataset


```{r}
sessionInfo() 
```

